---
title: "Simpson-Style Example of Context-Dependent Correlation"
author: "Kevin R. Coombes (analysis extended by Amir)"
date: "Original: 1 January 2026; Latest Run: `r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
rm(paths)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5, dev = "png",
                      fig.path = file.path(figdir, "R400"))
options(width=96)
.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
if (is.null(.format)) .format <- "broken"
.figger <- function(N, cap ) ifelse(.format != "latex",
                                 paste("Figure", N, ":",  cap),
                                 cap)
.taboo <- function(N, cap ) ifelse(.format %in% c("html", "docx"),
                                 paste("Table ", N, ": ",  cap, sep = ""),
                                 cap)
.sizzle <- function(pixels) ifelse(.format == "latex",
                                   pixels/2, pixels)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
We want a concrete Simpson-style example: a miR-mRNA pair whose *global* correlation
has the opposite sign of the *majority* of within-cohort correlations. We searched
the top 10 high-variance miRs and ranked pairs by spread plus sign reversal. The best
pair is **hsa-miR-200c-3p vs HTRA3** with global correlation **+0.403**, but **22**
cohorts are negative vs **9** positive (31 valid). The cohort spread is **1.178**
and median |r| is **0.272**; the most positive cohort is **UVM (r = 0.555)** and the
most negative is **READ (r = -0.623)**. We then visualize this in global and
cohort-specific plots, add a filtered Simpson-style view that removes low-|r|
cohorts, and quantify the heterogeneity with Cochran’s Q/I² plus a random-effects
meta-analysis of Fisher Z values.

# Background
The global (pan-cancer) correlation can be dominated by differences between cohorts.
To illustrate this clearly, we will identify a miR-mRNA pair with a strong correlation
across all samples, but an opposite pattern when we compute the correlations within
each cancer cohort.

# Data and Setup
Rationale: we load the matched miR/mRNA expression data and validate that cohort labels
and colors are available for plotting.
```{r load}
source("00-paths.R")
f1 <- file.path(paths$mirtcga, "matchedNoGBM.Rda")
f2 <- file.path(paths$mirtcga, "matchedNoGBM.rda")
if (file.exists(f1)) {
  load(f1)
} else if (file.exists(f2)) {
  load(f2)
} else {
  stop("Cannot find matchedNoGBM.Rda or matchedNoGBM.rda in ", paths$mirtcga)
}
rm(paths)
if (!exists("cohortColors")) {
  stop("cohortColors not found in matchedNoGBM data.")
}
if (!exists("colvec")) {
  colvec <- cohortColors[cancerType]
}
dim(miRdata)
dim(mRNAdata)
length(cancerType)
```
The matched data load cleanly. Next, we will confirm sample sizes per cohort to guide
thresholds for the local correlation analysis.

Rationale: we want to understand cohort sizes so we can avoid tiny cohorts when we
screen for sign flips.
```{r cohortSizes}
nPerCohort <- sort(table(cancerType), decreasing = TRUE)
nPerCohort
summary(as.vector(nPerCohort))
```
Most cohorts have large sample sizes, so we can safely require a minimum number of
cohorts with non-missing correlations. Next, we choose a small set of miRs to screen
based on variance across all samples.

# Search Strategy
Rationale: we start with miRs that vary the most across the full data set, because they
are more likely to produce strong, interpretable correlations.
```{r pickMir}
mirVar <- apply(miRdata, 1, var)
mirRank <- names(sort(mirVar, decreasing = TRUE))
topN <- 10
topMir <- mirRank[1:topN]
topMir
```
We now evaluate each of these miRs for a Simpson-style pattern by comparing global
correlations to the sign pattern across cohorts.

Rationale: for each candidate miR, we compute global correlations to all genes, then
compute within-cohort correlations and look for sign reversals with large spread.
```{r searchCandidates}
minCohorts <- 10
minOppCount <- 10
minMedianAbs <- 0.25
minGlobalAbs <- 0.20

candidateList <- list()
for (mir in topMir) {
  mAll <- miRdata[mir, ]
  global <- cor(t(mRNAdata), mAll)
  local <- sapply(levels(cancerType), function(ct) {
    idx <- which(cancerType == ct)
    cor(t(mRNAdata[, idx]), mAll[idx])
  })
  nValid <- rowSums(!is.na(local))
  nPos <- rowSums(local > 0, na.rm = TRUE)
  nNeg <- rowSums(local < 0, na.rm = TRUE)
  medAbs <- apply(local, 1, function(x) median(abs(x), na.rm = TRUE))
  spread <- apply(local, 1, function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

  oppCount <- ifelse(global > 0, nNeg, nPos)
  majOpp <- oppCount > (nValid / 2)
  flip <- !is.na(global) & nValid >= minCohorts &
    oppCount >= minOppCount & majOpp &
    medAbs >= minMedianAbs & abs(global) >= minGlobalAbs

  if (any(flip)) {
    score <- spread * abs(global)
    score[!flip] <- NA
    pick <- which.max(score)
    candidateList[[mir]] <- data.frame(
      miR = mir,
      gene = rownames(mRNAdata)[pick],
      global = global[pick],
      spread = spread[pick],
      nPos = nPos[pick],
      nNeg = nNeg[pick],
      nValid = nValid[pick],
      medAbs = medAbs[pick],
      score = score[pick],
      stringsAsFactors = FALSE
    )
  }
}

candidates <- do.call(rbind, candidateList)
candidates <- candidates[order(-candidates$score), ]
head(candidates, 5)
```
We identified one or more strong candidates. Next, we select the top-ranked pair and
compute the full cohort-wise correlation profile for visualization.

# Detailed Look at the Best Pair
Rationale: we recompute cohort correlations for the top-ranked pair and extract the
most extreme positive and negative cohorts for focused plots.
```{r bestPair}
best <- candidates[1, ]
bestMir <- best$miR
bestGene <- best$gene

mAll <- miRdata[bestMir, ]
globalBest <- cor(mRNAdata[bestGene, ], mAll)
localBest <- sapply(levels(cancerType), function(ct) {
  idx <- which(cancerType == ct)
  cor(mRNAdata[bestGene, idx], mAll[idx])
})
nPerCohort <- table(cancerType)
localTab <- data.frame(
  Cohort = names(localBest),
  Correlation = as.vector(localBest),
  N = as.vector(nPerCohort[names(localBest)]),
  stringsAsFactors = FALSE
)
localTab <- localTab[order(localTab$Correlation, decreasing = TRUE), ]
localTab
```
The top-ranked pair shows clear sign disagreement: the global correlation has the
opposite sign from most cohorts. Now we will visualize this in three steps: (1) global
scatter, (2) cohort correlation bars, and (3) two contrasting cohorts.

Rationale: the global scatter plot shows the overall correlation across all samples.
```{r figGlobal, fig.cap=.figger(1, "Global miR-mRNA correlation across all cohorts."), fig.width=7, fig.height=6}
plot(mRNAdata[bestGene, ], mAll, col = colvec, pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = paste("Global correlation:", round(globalBest, 3)))
```
The global scatter highlights a clear overall trend. Next, we compare it to the
cohort-by-cohort correlations to show the contradiction explicitly.

Rationale: a barplot of cohort correlations makes the sign reversal easy to read.
```{r figBars, fig.cap=.figger(2, "Within-cohort correlations for the Simpson-style pair."), fig.width=10, fig.height=6}
opar <- par(mar = c(8, 4, 2, 1))
pts <- barplot(localTab$Correlation, names.arg = localTab$Cohort,
               col = cohortColors[localTab$Cohort], las = 2, cex.names = 0.7,
               ylab = "Correlation", main = "Cohort correlations (sorted)")
abline(h = globalBest, lwd = 2, col = "black")
par(opar)
```
The barplot shows the mismatch between the global estimate and the local pattern.
Next, we inspect one positive and one negative cohort in detail.

Rationale: contrasting two cohorts makes the Simpson-style reversal concrete in raw
expression space.
```{r figCohorts, fig.cap=.figger(3, "Contrasting cohorts with opposite-sign correlations."), fig.width=10, fig.height=5}
posCohort <- localTab$Cohort[1]
negCohort <- localTab$Cohort[nrow(localTab)]

par(mfrow = c(1,2))
idxPos <- which(cancerType == posCohort)
plot(mRNAdata[bestGene, idxPos], mAll[idxPos], col = cohortColors[posCohort], pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = paste(posCohort, "corr =", round(localBest[posCohort], 3)))

idxNeg <- which(cancerType == negCohort)
plot(mRNAdata[bestGene, idxNeg], mAll[idxNeg], col = cohortColors[negCohort], pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = paste(negCohort, "corr =", round(localBest[negCohort], 3)))
par(mfrow = c(1,1))
```
The two cohort plots show opposite slopes, confirming that the global correlation is
not representative of most cohort-level relationships. Next, we will report a short
table of sign counts to make the logic explicit.

Rationale: a compact table summarizes the sign distribution across cohorts.
```{r signTable}
signSummary <- data.frame(
  Global = globalBest,
  N_Pos = sum(localBest > 0, na.rm = TRUE),
  N_Neg = sum(localBest < 0, na.rm = TRUE),
  N_Valid = sum(!is.na(localBest)),
  Median_Abs = median(abs(localBest), na.rm = TRUE)
)
signSummary
```
This summary confirms that the global sign is opposite to the majority of cohort
signs, which is the core Simpson-style effect we wanted. At this point, we have a
concrete example and three linked figures suitable for a manuscript or supplement.

# Decomposition: Between vs Within Cohorts
Rationale: quantify how much of the global correlation is driven by between-cohort
differences versus within-cohort covariance.
```{r decomp}
coh <- as.character(cancerType)
mx <- tapply(mRNAdata[bestGene, ], coh, mean)
my <- tapply(mAll, coh, mean)

N <- length(mAll)
muX <- mean(mRNAdata[bestGene, ])
muY <- mean(mAll)

covGlobal <- cov(mRNAdata[bestGene, ], mAll)
covWithin <- 0
covBetween <- 0
for (ct in names(mx)) {
  idx <- which(cancerType == ct)
  nx <- length(idx)
  if (nx < 2) next
  cx <- mRNAdata[bestGene, idx]
  cy <- mAll[idx]
  covWithin <- covWithin + (nx - 1) * cov(cx, cy)
  covBetween <- covBetween + nx * (mean(cx) - muX) * (mean(cy) - muY)
}
covWithin <- covWithin/(N - 1)
covBetween <- covBetween/(N - 1)

data.frame(Global = covGlobal, Within = covWithin, Between = covBetween)
```
The between-cohort covariance is strongly positive, while the within-cohort
covariance is negative. This is the quantitative driver of the Simpson reversal.

# Heterogeneity Test (Cochran’s Q, I²)
Rationale: test whether the cohort-specific correlations are statistically
consistent with a single common correlation. The null hypothesis is that all
cohorts share the same true correlation. The p-value is the chi-squared test for
heterogeneity.
```{r heterogeneity}
nPer <- table(cancerType)
validCoh <- names(localBest)[!is.na(localBest)]
nCoh <- as.numeric(nPer[validCoh])
z <- atanh(localBest[validCoh])
w <- nCoh - 3
zbar <- sum(w * z)/sum(w)

Q <- sum(w * (z - zbar)^2)
df <- length(z) - 1
pval <- 1 - pchisq(Q, df)
I2 <- max(0, (Q - df)/Q)

data.frame(Q = Q, df = df, p_value = pval, I2 = I2)
```
The Q test evaluates heterogeneity of Fisher Z across cohorts. A small p-value
indicates that the cohort correlations are not equal, supporting context dependence.

# Random-Effects Meta-Analysis of Correlations
Rationale: compute a pooled correlation that allows between-cohort variance (tau²)
using a standard DerSimonian–Laird random-effects model on Fisher Z values.
```{r randomEffects}
denom <- sum(w) - sum(w^2)/sum(w)
tau2 <- max(0, (Q - df)/denom)

wRE <- 1/(1/w + tau2)
zRE <- sum(wRE * z)/sum(wRE)
rRE <- (exp(2 * zRE) - 1)/(exp(2 * zRE) + 1)

data.frame(tau2 = tau2, zRE = zRE, rRE = rRE)
```
The random-effects estimate reflects the average cohort correlation while accounting
for between-cohort heterogeneity.

# Canonical Simpson Visualization (Cohort Means)
Rationale: show the cohort-mean trend directly (Simpson-style).
```{r figMeans, fig.cap=.figger(5, "Cohort means for the Simpson pair."), fig.width=7, fig.height=5}
plot(mx, my, col = cohortColors[names(mx)], pch = 16,
     xlab = paste(bestGene, "(mean)"), ylab = paste(bestMir, "(mean)"),
     main = paste("Cohort means r =", round(cor(mx, my), 3)))
abline(lm(my ~ mx), lwd = 2)
text(mx, my, labels = names(mx), col = cohortColors[names(mx)],
     pos = 3, cex = 0.7)
```
The cohort means show a strong positive trend, even though most within-cohort
correlations are negative.


# Filtered Simpson-Style Visualization
Rationale: to sharpen the visual contrast, we keep only the top K cohorts by |r|
and overlay the global trend against per-cohort trends for those groups.
```{r filterCohorts}
K <- 5
ord <- order(abs(localTab$Correlation), decreasing = TRUE)
keep <- localTab$Cohort[ord[1:K]]
keep
```
These cohorts have the strongest within-cohort signals and are best suited for a
clean Simpson-style visualization. Next we plot all points in light gray, then
overlay the selected cohorts with their own regression lines and the global line.

Rationale: the global line should slope one way, while most cohort lines slope the
other way.
```{r figSimpson, fig.cap=.figger(4, "Simpson-style visualization using filtered cohorts. Gray = all samples, colored = selected cohorts; black line = global trend; colored lines = within-cohort trends."), fig.width=8, fig.height=6}
plot(mRNAdata[bestGene, ], mAll, col = "gray80", pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = paste(bestMir, "vs", bestGene))
abline(lm(mAll ~ mRNAdata[bestGene, ]), lwd = 2)

for (ct in keep) {
  idx <- which(cancerType == ct)
  points(mRNAdata[bestGene, idx], mAll[idx], col = cohortColors[ct], pch = 16)
  abline(lm(mAll[idx] ~ mRNAdata[bestGene, idx]), col = cohortColors[ct], lwd = 2)
}
```
The global line (black) now contrasts clearly with the majority of cohort lines,
which highlights the Simpson-style reversal in a single panel.

# Four-Panel Summary Figure
Rationale: combine the key plots into a single multi-panel figure for the main text.
```{r figPanel, fig.cap=.figger(6, "Four-panel summary of Simpson-style behavior."), fig.width=12, fig.height=10}
opar <- par(mfrow = c(2, 2))

# Panel A: global scatter
plot(mRNAdata[bestGene, ], mAll, col = "gray80", pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = paste("Global r =", round(globalBest, 3)))
abline(lm(mAll ~ mRNAdata[bestGene, ]), lwd = 2)

# Panel B: cohort correlations (barplot)
par(mar = c(8, 4, 2, 1))
barplot(localTab$Correlation, names.arg = localTab$Cohort,
        col = cohortColors[localTab$Cohort], las = 2, cex.names = 0.7,
        ylab = "Correlation", main = "Cohort correlations")
abline(h = globalBest, lwd = 2, col = "black")

# Panel C: cohort means
par(mar = c(5, 4, 2, 1))
plot(mx, my, col = cohortColors[names(mx)], pch = 16,
     xlab = paste(bestGene, "(mean)"), ylab = paste(bestMir, "(mean)"),
     main = paste("Cohort means r =", round(cor(mx, my), 3)))
abline(lm(my ~ mx), lwd = 2)

# Panel D: filtered Simpson plot
plot(mRNAdata[bestGene, ], mAll, col = "gray80", pch = 16,
     xlab = bestGene, ylab = bestMir,
     main = "Filtered cohorts")
abline(lm(mAll ~ mRNAdata[bestGene, ]), lwd = 2)
for (ct in keep) {
  idx <- which(cancerType == ct)
  points(mRNAdata[bestGene, idx], mAll[idx], col = cohortColors[ct], pch = 16)
  abline(lm(mAll[idx] ~ mRNAdata[bestGene, idx]), col = cohortColors[ct], lwd = 2)
}

par(opar)
```
This combined figure captures the global trend, cohort variability, mean-level
Simpson structure, and the filtered reversal in one place.

# Conclusions
We isolated a single miR-mRNA pair where the global correlation reverses the direction
of the majority of cohort-specific correlations. This provides a concrete and visual
illustration of why subgroup structure can invalidate naive interpretations of
pan-cancer correlation.
