---
title: "Case Studies: Context-Dependent miR Behavior Across Cancer Types"
author: "Amir (Building on Kevin R. Coombes' framework)"
date: "Original: `r Sys.Date()`; Latest Run: `r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5, dev = "png",
                      fig.path = file.path(figdir, "R402"))
options(width=96)
.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
if (is.null(.format)) .format <- "broken"
.figger <- function(N, cap ) ifelse(.format != "latex",
                                 paste("Figure", N, ":",  cap),
                                 cap)
.taboo <- function(N, cap ) ifelse(.format %in% c("html", "docx"),
                                 paste("Table ", N, ": ",  cap, sep = ""),
                                 cap)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary

**What we learned from the pathway analysis (401):**
In our previous analysis, we discovered that context-dependency is the *rule*, not
the exception. For well-studied miRNAs like let-7a, miR-34a, and miR-362, essentially
100% of genes show context-dependent (mixed positive/negative) correlation across
cancer types. Even for miR-21 (the most studied oncomiR), only 14 out of 20,193
genes (0.07%) show consistently negative correlation.

**Goal of this analysis:**
Dive deeper into specific case studies to understand *why* certain miR-gene pairs
behave differently across cancer types. We will:

1. Examine let-7a (tumor suppressor) - which cancers show strongest negative correlation?
2. Study miR-34a (p53-regulated) - does it behave as a tumor suppressor uniformly?
3. Extend Kevin's miR-362 example with additional functional analysis
4. Identify cancer types that cluster by miR correlation patterns

**Key Questions:**
- Are there cancer-specific miR "signatures" in correlation patterns?
- Do known oncogenic contexts show weaker tumor suppressor correlations?
- Can we identify tissue-specific miR targets?

# Getting Started

```{r paths, message=FALSE, warning=FALSE}
source("00-paths.R")

library(ggplot2)
library(pheatmap)
library(RColorBrewer)
```

```{r loaddata}
load(file.path(paths$mirtcga, "matchedNoGBM.rda"))
load(file.path(paths$mirtcga, "cohortColors.Rda"))

cat("Data loaded:\n")
cat("  Genes:", nrow(mRNAdata), "\n")
cat("  miRs:", nrow(miRdata), "\n")
cat("  Samples:", ncol(mRNAdata), "\n")
cat("  Cohorts:", paste(sort(unique(cancerType)), collapse=", "), "\n")
```

# Utility Functions

I'll use the same correlation approach as before, but add functions for
visualization and comparison across cohorts.

```{r utilities}
# Compute correlation for one miR with all genes in one cohort
compute_cohort_cor <- function(mir_name, cohort, mRNAdata, miRdata, cancerType) {
  idx <- cancerType == cohort
  if (sum(idx) < 10) return(rep(NA, nrow(mRNAdata)))

  mir_expr <- as.numeric(miRdata[mir_name, idx])
  if (all(mir_expr == 0) || sd(mir_expr) == 0) return(rep(NA, nrow(mRNAdata)))

  cors <- sapply(1:nrow(mRNAdata), function(i) {
    mrna_expr <- as.numeric(mRNAdata[i, idx])
    if (sd(mrna_expr) > 0) {
      cor(mir_expr, mrna_expr, use = "complete.obs")
    } else {
      NA
    }
  })
  names(cors) <- rownames(mRNAdata)
  return(cors)
}

# Compute correlation matrix for one miR across all cohorts
compute_mir_matrix <- function(mir_name, mRNAdata, miRdata, cancerType) {
  cohorts <- sort(unique(as.character(cancerType)))
  result <- matrix(NA, nrow = nrow(mRNAdata), ncol = length(cohorts),
                   dimnames = list(rownames(mRNAdata), cohorts))

  for (ct in cohorts) {
    result[, ct] <- compute_cohort_cor(mir_name, ct, mRNAdata, miRdata, cancerType)
  }
  return(result)
}
```

# Case Study 1: let-7a - The Canonical Tumor Suppressor

Let-7a was one of the first discovered miRNAs and is well-established as a tumor
suppressor. It directly targets oncogenes like RAS and HMGA2. If correlation reflects
regulation, we should see consistent negative correlation with these targets.

```{r let7a_setup}
# Find let-7a variants
let7a_mirs <- grep("let-7a", rownames(miRdata), value = TRUE)
cat("Available let-7a variants:\n")
print(let7a_mirs)

# Use the 5p form (more commonly studied)
let7a <- grep("let-7a-5p", rownames(miRdata), value = TRUE)[1]
cat("\nUsing:", let7a, "\n")
```

```{r let7a_compute}
cat("Computing correlations for", let7a, "across all cohorts...\n")
let7a_cors <- compute_mir_matrix(let7a, mRNAdata, miRdata, cancerType)
cat("Done!\n")
```

Let me first look at the distribution of correlations across all cohorts.

```{r let7a_dist, fig.cap="Distribution of let-7a correlations by cancer type.", fig.width=12, fig.height=6}
# Calculate summary statistics per cohort
let7a_summary <- data.frame(
  cohort = colnames(let7a_cors),
  mean_cor = apply(let7a_cors, 2, mean, na.rm = TRUE),
  median_cor = apply(let7a_cors, 2, median, na.rm = TRUE),
  n_negative = apply(let7a_cors, 2, function(x) sum(x < -0.3, na.rm = TRUE)),
  n_positive = apply(let7a_cors, 2, function(x) sum(x > 0.3, na.rm = TRUE))
)

# Which cohorts have the most negative correlations?
cat("Cohorts ranked by genes with strong negative correlation (r < -0.3):\n")
print(let7a_summary[order(-let7a_summary$n_negative), c("cohort", "n_negative", "n_positive")])
```

This is fascinating! Let's visualize which cohorts show the strongest tumor
suppressor-like behavior (negative correlations).

```{r let7a_barplot, fig.cap="Genes with strong let-7a correlation by cancer type.", fig.width=10, fig.height=6}
# Melt for plotting
let7a_plot <- let7a_summary[order(let7a_summary$n_negative), ]
let7a_plot$cohort <- factor(let7a_plot$cohort, levels = let7a_plot$cohort)

par(mfrow = c(1, 2))
barplot(let7a_plot$n_negative, names.arg = let7a_plot$cohort, las = 2,
        col = cohortColors[as.character(let7a_plot$cohort)],
        main = "Genes with r < -0.3 (let-7a)", ylab = "Number of genes")

barplot(let7a_plot$n_positive, names.arg = let7a_plot$cohort, las = 2,
        col = cohortColors[as.character(let7a_plot$cohort)],
        main = "Genes with r > 0.3 (let-7a)", ylab = "Number of genes")
```

## Known let-7a Targets

Let's check if known oncogenic targets of let-7a show negative correlation.
Classic let-7a targets include RAS genes, HMGA2, and MYC.

```{r let7a_targets}
# Known let-7a targets
known_targets <- c("KRAS", "NRAS", "HRAS", "HMGA2", "MYC", "LIN28A", "LIN28B", "IGF2BP1", "IGF2BP2")

# Check which are in our data
available_targets <- known_targets[known_targets %in% rownames(let7a_cors)]
cat("Available known targets:", paste(available_targets, collapse = ", "), "\n\n")

# Get their correlations across cohorts
if (length(available_targets) > 0) {
  target_cors <- let7a_cors[available_targets, , drop = FALSE]

  cat("Correlation of known let-7a targets across cohorts:\n")
  print(round(target_cors, 3))

  cat("\nSummary: proportion of cohorts with negative correlation:\n")
  prop_neg <- apply(target_cors, 1, function(x) mean(x < 0, na.rm = TRUE))
  print(round(prop_neg, 2))
}
```

```{r let7a_targets_heatmap, fig.cap="Correlation of known let-7a targets across cancer types.", fig.width=12, fig.height=5}
if (length(available_targets) > 2) {
  # Remove NA rows/cols
  target_cors_clean <- target_cors[, colSums(!is.na(target_cors)) > 0]

  pheatmap(target_cors_clean,
           main = "let-7a correlation with known targets",
           color = colorRampPalette(c("blue", "white", "red"))(100),
           breaks = seq(-1, 1, length.out = 101),
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           fontsize_row = 10,
           fontsize_col = 8)
}
```

Interesting! Even for *known* direct targets of let-7a, the correlation patterns
are highly variable across cancer types. This strongly supports Kevin's thesis
that context matters enormously.

# Case Study 2: miR-34a - The p53 Effector

miR-34a is transcriptionally activated by p53 and mediates tumor suppression.
Its targets include CDK6, E2F3, and BCL2. Since p53 status varies across cancers,
we might expect miR-34a to behave differently depending on p53 functionality.

```{r mir34a_setup}
mir34a <- grep("miR-34a-5p", rownames(miRdata), value = TRUE)[1]
cat("Analyzing:", mir34a, "\n")

cat("Computing correlations...\n")
mir34a_cors <- compute_mir_matrix(mir34a, mRNAdata, miRdata, cancerType)
cat("Done!\n")
```

```{r mir34a_summary}
# Summary statistics
mir34a_summary <- data.frame(
  cohort = colnames(mir34a_cors),
  mean_cor = apply(mir34a_cors, 2, mean, na.rm = TRUE),
  n_strong_neg = apply(mir34a_cors, 2, function(x) sum(x < -0.3, na.rm = TRUE)),
  n_strong_pos = apply(mir34a_cors, 2, function(x) sum(x > 0.3, na.rm = TRUE))
)

cat("miR-34a: Cohorts with most strongly correlated genes:\n")
print(mir34a_summary[order(-mir34a_summary$n_strong_neg), ])
```

## Known miR-34a Targets

```{r mir34a_targets}
# Known miR-34a targets (cell cycle, apoptosis)
mir34a_known <- c("CDK6", "CDK4", "E2F3", "BCL2", "NOTCH1", "MET", "SIRT1", "MYC")
available_34a <- mir34a_known[mir34a_known %in% rownames(mir34a_cors)]

cat("Checking known miR-34a targets:\n")
if (length(available_34a) > 0) {
  target34_cors <- mir34a_cors[available_34a, , drop = FALSE]
  print(round(target34_cors, 3))

  cat("\nProportion of cohorts with negative correlation:\n")
  print(round(apply(target34_cors, 1, function(x) mean(x < 0, na.rm = TRUE)), 2))
}
```

# Case Study 3: Deep Dive into miR-362 (Kevin's Example)

Kevin showed that miR-362 has Cohen's kappa of 0.120 in PRAD but only 0.003 in SKCM
for association with validated targets. Let me explore this more.

```{r mir362_setup}
mir362 <- grep("miR-362-3p", rownames(miRdata), value = TRUE)[1]
cat("Analyzing:", mir362, "\n")

mir362_cors <- compute_mir_matrix(mir362, mRNAdata, miRdata, cancerType)
```

```{r mir362_compare, fig.cap="miR-362 correlation patterns in PRAD vs SKCM.", fig.width=10, fig.height=5}
# Compare PRAD and SKCM directly
par(mfrow = c(1, 2))

# PRAD
hist(mir362_cors[, "PRAD"], breaks = 50, main = "miR-362 in PRAD",
     xlab = "Correlation", col = cohortColors["PRAD"])
abline(v = 0, col = "red", lwd = 2, lty = 2)

# SKCM
hist(mir362_cors[, "SKCM"], breaks = 50, main = "miR-362 in SKCM",
     xlab = "Correlation", col = cohortColors["SKCM"])
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

```{r mir362_scatter, fig.cap="Comparing miR-362 correlations: PRAD vs SKCM."}
# Scatter plot comparing the two
plot(mir362_cors[, "PRAD"], mir362_cors[, "SKCM"],
     pch = 16, col = rgb(0, 0, 0, 0.1), cex = 0.5,
     xlab = "Correlation in PRAD", ylab = "Correlation in SKCM",
     main = "miR-362: PRAD vs SKCM (each point = 1 gene)")
abline(0, 1, col = "red", lwd = 2)
abline(h = 0, v = 0, col = "gray", lty = 2)

# Correlation between the correlation profiles
rho <- cor(mir362_cors[, "PRAD"], mir362_cors[, "SKCM"], use = "complete.obs")
legend("topleft", legend = paste("r =", round(rho, 3)), bty = "n", cex = 1.2)
```

The correlation between PRAD and SKCM correlation profiles is only `r round(rho, 3)`!
This means the same miR has essentially *unrelated* correlation patterns in these
two cancer types.

## Genes with Opposite Behavior

Let's identify genes where miR-362 shows opposite correlation in PRAD vs SKCM.

```{r mir362_opposite}
# Find genes that flip sign
prad_neg_skcm_pos <- which(mir362_cors[, "PRAD"] < -0.2 & mir362_cors[, "SKCM"] > 0.2)
prad_pos_skcm_neg <- which(mir362_cors[, "PRAD"] > 0.2 & mir362_cors[, "SKCM"] < -0.2)

cat("Genes with opposite correlation patterns:\n")
cat("  Negative in PRAD, Positive in SKCM:", length(prad_neg_skcm_pos), "\n")
cat("  Positive in PRAD, Negative in SKCM:", length(prad_pos_skcm_neg), "\n")

if (length(prad_neg_skcm_pos) > 0) {
  cat("\nTop genes (negative PRAD, positive SKCM):\n")
  flip_genes <- names(prad_neg_skcm_pos)
  flip_data <- data.frame(
    gene = flip_genes,
    PRAD = mir362_cors[flip_genes, "PRAD"],
    SKCM = mir362_cors[flip_genes, "SKCM"]
  )
  print(head(flip_data[order(flip_data$PRAD - flip_data$SKCM), ], 10))
}
```

# Clustering Cancer Types by miR Correlation Patterns

Do cancer types cluster based on how miRNAs correlate with mRNAs? This could
reveal tissue-level regulatory patterns.

```{r cluster_cohorts, fig.cap="Hierarchical clustering of cancer types by miR-21 correlation patterns.", fig.width=10, fig.height=8}
# Use miR-21 as an example (well-expressed across all cohorts)
mir21 <- grep("miR-21-5p", rownames(miRdata), value = TRUE)[1]
mir21_cors <- compute_mir_matrix(mir21, mRNAdata, miRdata, cancerType)

# Select top 1000 most variable genes for clustering
gene_var <- apply(mir21_cors, 1, var, na.rm = TRUE)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:1000]
mir21_top <- mir21_cors[top_genes, ]

# Remove cohorts with too many NAs
valid_cohorts <- colSums(!is.na(mir21_top)) > 500
mir21_top <- mir21_top[, valid_cohorts]

# Replace remaining NAs with 0 for clustering
mir21_top[is.na(mir21_top)] <- 0

# Cluster cohorts
pheatmap(t(mir21_top),
         main = "Cancer cohorts clustered by miR-21 correlation patterns\n(top 1000 most variable genes)",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-0.8, 0.8, length.out = 101),
         show_colnames = FALSE,
         fontsize_row = 8,
         clustering_method = "ward.D2")
```

# Summary of Findings

This case study analysis reveals several key insights:

1. **Known targets don't show consistent correlation**: Even for well-established
   direct targets of tumor suppressor miRNAs (let-7a, miR-34a), the correlation
   patterns are highly variable across cancer types.

2. **Same miR, different behavior**: miR-362 has essentially uncorrelated gene
   correlation profiles in PRAD vs SKCM (r ≈ 0). Many genes show *opposite*
   correlations in these two cancers.

3. **Cancer types cluster by miR behavior**: When we cluster cancers by their
   miR-gene correlation patterns, we see tissue-specific groupings, suggesting
   that the cellular context fundamentally shapes miR-mRNA relationships.

4. **Implications**: Researchers should be extremely cautious about:
   - Assuming pan-cancer miR target predictions apply to specific cancer types
   - Using global correlation to infer regulatory relationships
   - Ignoring the tissue context when studying miR function

These findings provide additional support for Kevin's central thesis that
"correlation is not regulation" and that subgroup structure must be considered
in correlation analyses.

# Appendix
```{r si}
sessionInfo()
```
