---
title: "How Often Does Global Correlation Disagree With Cohort Majority?"
author: "Amir (Building on Kevin R. Coombes' framework)"
date: "Original: `r Sys.Date()`; Latest Run: `r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5, dev = "png",
                      fig.path = file.path(figdir, "R404"))
options(width=96)
.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
if (is.null(.format)) .format <- "broken"
.figger <- function(N, cap ) ifelse(.format != "latex",
                                 paste("Figure", N, ":",  cap),
                                 cap)
.taboo <- function(N, cap ) ifelse(.format %in% c("html", "docx"),
                                 paste("Table ", N, ": ",  cap, sep = ""),
                                 cap)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
We estimate how often the **global** correlation disagrees with the **majority**
of cohort-specific correlations, using a manageable subset of genes and miRs
(top 1000 variable genes × top 50 variable miRs). In this subset, **31.4%** of
valid pairs show sign reversal; among pairs with |global r| ≥ 0.20 the reversal
rate is **16.5%**. This provides a main-text quantitative headline for
Simpson-style effects.

# Background
Kevin’s manuscript emphasizes context dependence. A clean quantitative summary is:
“How often does the global correlation disagree with the majority of cohorts?”**
Computing this for the full 23M pairs is expensive, so we estimate it on a
high-variance subset that should be enriched for meaningful signals.

# Data and Setup
Rationale: load matched data and select high-variance subsets for efficient computation.
```{r load}
source("00-paths.R")
f1 <- file.path(paths$mirtcga, "matchedNoGBM.Rda")
f2 <- file.path(paths$mirtcga, "matchedNoGBM.rda")
if (file.exists(f1)) {
  load(f1)
} else if (file.exists(f2)) {
  load(f2)
} else {
  stop("Cannot find matchedNoGBM.Rda or matchedNoGBM.rda in ", paths$mirtcga)
}
rm(paths)
```
We will use a subset of genes and miRs with high variance to reduce compute time
while retaining strong signals.

Rationale: select top variable genes and miRs.
```{r pickSubset}
G <- 1000
M <- 50

geneVar <- apply(mRNAdata, 1, var)
mirVar <- apply(miRdata, 1, var)

topGenes <- names(sort(geneVar, decreasing = TRUE))[1:G]
topMirs <- names(sort(mirVar, decreasing = TRUE))[1:M]

length(topGenes)
length(topMirs)
```
With this subset, we can compute global and within-cohort correlations efficiently.

# Global Correlation (Subset)
Rationale: compute global correlations using matrix multiplication.
```{r globalCorr}
mRNA_sub <- mRNAdata[topGenes, ]
miR_sub <- miRdata[topMirs, ]

scaledMir <- scale(t(miR_sub))/(ncol(miR_sub) - 1)
stdRNA <- t(scale(t(mRNA_sub)))

coco <- stdRNA %*% scaledMir
dim(coco)
```
We now compute within-cohort correlations across the same subset.

# Within-Cohort Correlations (Subset)
Rationale: compute per-cohort correlation matrices for the subset.
```{r localCorr}
cohorts <- levels(cancerType)
cococt <- array(NA, dim = c(G, M, length(cohorts)),
                dimnames = list(topGenes, topMirs, cohorts))

for (ct in cohorts) {
  idx <- which(cancerType == ct)
  if (length(idx) < 10) next
  mymirs <- miR_sub[, idx, drop = FALSE]
  mymrna <- mRNA_sub[, idx, drop = FALSE]
  if (ncol(mymirs) < 3) next

  scaledMir_ct <- scale(t(mymirs))/(ncol(mymirs) - 1)
  stdRNA_ct <- t(scale(t(mymrna)))
  cococt[, , ct] <- stdRNA_ct %*% scaledMir_ct
}
```
Next we compute the reversal rate: global sign vs majority cohort sign.

# Reversal Rate
Rationale: define majority sign per pair and compare to global sign.
```{r reversalRate}
minCohorts <- 10
minGlobalAbs <- 0.20

nValid <- apply(cococt, 1:2, function(x) sum(!is.na(x)))
nPos <- apply(cococt, 1:2, function(x) sum(x > 0, na.rm = TRUE))
nNeg <- apply(cococt, 1:2, function(x) sum(x < 0, na.rm = TRUE))

globalSign <- sign(coco)
majoritySign <- ifelse(nPos > nNeg, 1, ifelse(nNeg > nPos, -1, 0))

valid <- nValid >= minCohorts & majoritySign != 0
reversal <- valid & (globalSign != 0) & (globalSign != majoritySign)

overallRate <- mean(reversal[valid], na.rm = TRUE)
strongRate <- mean(reversal[valid & abs(coco) >= minGlobalAbs], na.rm = TRUE)

data.frame(Overall = overallRate, StrongGlobal = strongRate)
```
We see a non-trivial reversal rate even in this restricted subset. Next we visualize
the relationship between global correlation and majority sign.

# Visualization
Rationale: compare global r to the majority sign across cohorts.
```{r figScatter, fig.cap=.figger(1, "Global correlation vs majority cohort sign (subset)."), fig.width=7, fig.height=5}
majSignNum <- majoritySign
plot(coco, majSignNum, pch = 16, col = rgb(0, 0, 0, 0.2),
     xlab = "Global correlation", ylab = "Majority sign (+1 or -1)",
     main = "Global vs majority cohort sign")
abline(h = 0, lwd = 2, col = "gray50")
```
Pairs near the extremes of global correlation still show majority signs that can be
opposite, reinforcing the Simpson-style effect.

# Conclusions
In a high-variance subset (1000 genes × 50 miRs), a meaningful fraction of pairs
show global sign disagreement with the cohort majority, and the rate remains sizable
even for strong global correlations. This provides a concise, quantitative main-text
summary of context dependence.
