---
title: "Robustness: Effect-size–conditioned mixed-sign + Simpson reversals"
author: "Amir (major revision; toolkit framing)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
rm(paths)
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
We want a reviewer-proof answer to: “Is mixed-sign / Simpson reversal just noise near zero?”

We compute effect-size–conditioned rates on a **large, high-variance domain**
(top 5,000 genes × top 200 miRNAs = 1,000,000 pairs), and we use a tolerance
**ε ∈ {0, 0.05, 0.10}** to ignore tiny correlations when counting signs.

Key headline (ε=0.05):

- Mixed-sign remains **~99.5%** even when conditioning on `max_c |r_c| ≥ 0.2`.
- Simpson reversal remains **~13.3%** among pairs with `|r_global| ≥ 0.2`.

The script writes a manuscript-ready figure:
`doc/bioinformatics/figures/Figure_RobustnessSignFlip.pdf`.

# Why this analysis
The original “mixed sign across cohorts” claim can be dismissed as an artifact of tiny correlations flipping sign.

Rationale: we explicitly (1) ignore near-zero within-cohort correlations, and (2) condition on effect size before asking for sign disagreement.

# Run the robustness script
Rationale: use a single script to keep the manuscript numbers consistent and reproducible.
```{r run, message=TRUE}
system("Rscript scripts/robustness_signflip.R", intern = FALSE)
```

# Inspect the summary table
Rationale: load the CSV and confirm the key thresholds used in the manuscript (t = 0.2, ε = 0.05).
```{r inspect}
tab <- read.csv("results/robustness_signflip_summary.csv")
subset(tab, eps == 0.05 & threshold %in% c(0, 0.2, 0.3))
```

Interpretation: the rates do not collapse when we require non-trivial correlations; reversals remain common even among “strong” pooled correlations.

