---
title: "Standard heterogeneity reporting: Cochran’s Q, I², and FDR"
author: "Amir (major revision; toolkit framing)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
rm(paths)
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
We replace the old, nonstandard “χ²=60/100 bins” with standard meta-analysis reporting:

- Cochran’s **Q** p-value
- **I²** (fraction of variability due to true heterogeneity)
- **BH-FDR** across the tested domain

We compute these on the same high-variance domain used for robustness:
top 5,000 genes × top 200 miRNAs = 1,000,000 pairs.

Key headline used in the manuscript:

- median **I² ≈ 0.86** (IQR ≈ 0.80–0.90)
- **~99.5%** of pairs show heterogeneity at FDR < 0.05

# Why this analysis
Rationale: fixed χ² cutoffs are not comparable across varying degrees of freedom and are not standard in meta-analysis. Q/I² are standard and interpretable.

# Run the heterogeneity script
Rationale: compute and cache Q/I²/FDR summaries + write a manuscript-ready figure.
```{r run, message=TRUE}
system("Rscript scripts/heterogeneity_I2_FDR.R", intern = FALSE)
```

# Inspect the summary table
Rationale: confirm the reported median I² and FDR fraction.
```{r inspect}
summ <- read.csv("results/heterogeneity_summary.csv")
summ
```

Interpretation: heterogeneity is not a niche phenomenon; it is widespread and large even after focusing on a tractable but signal-rich domain.

