---
title: "Sensitivity: Pearson vs Spearman"
author: "Amir (major revision; toolkit framing)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r sigh, echo=FALSE}
source("00-paths.R")
figdir <- file.path(paths$results, "Figures")
if (!dir.exists(figdir)) dir.create(figdir, recursive = TRUE)
rm(paths)
```

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
Reviewer-friendly sensitivity check: do our main qualitative conclusions depend on Pearson correlation?

We compute:

1) Concordance between global Pearson r and global Spearman ρ across a high-variance subset.
2) Simpson reversal rate under Spearman (same ε-based sign logic as the main robustness analysis).

Headline:

- Pearson vs Spearman concordance is **~0.94**.
- Simpson reversal under Spearman is **~12.6%** among pairs with `|ρ_global| ≥ 0.2` (ε=0.05).

# Run the sensitivity script
Rationale: keep the check minimal and reproducible.
```{r run, message=TRUE}
system("Rscript scripts/spearman_sensitivity.R", intern = FALSE)
```

# Inspect outputs
Rationale: load the CSV and confirm the numbers used in the supplement caption.
```{r inspect}
read.csv("results/spearman_sensitivity.csv")
```

Interpretation: the findings are not an artifact of Pearson correlation or outliers; Spearman yields a similar qualitative picture.

