---
title: "Main Figure 1: Global correlation distribution + miR-122 cohort effects"
author: "Amir (major revision; toolkit framing)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Summary
We want an early, intuitive figure for the main text that sets expectations:

1) Global (pooled) correlations span a wide range.
2) Extreme correlations can be explained by between-cohort expression differences.

This report regenerates the manuscript’s Figure 1:
`doc/bioinformatics/figures/Figure_GlobalDistribution.pdf`.

# Generate Figure 1
Rationale: build the figure through a single script so the LaTeX manuscript and the report stay in sync.
```{r run, message=TRUE}
system("Rscript scripts/figure_global_distribution.R", intern = FALSE)
```

