---
title: "TCGA Cohorts"
author: "Kevin R. Coombes"
date: "8 August 2018"
output: 
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Background
The Cancer Genome Atlas (TCGA) divides the data by cancer type, which is referred to
at most of the relevant web sites (or data sources) as a "_cohort_". The goal of this
report is simply to get a list of those cohorts.

# Cohorts
First, we define the paths that we will use to store things.
```{r paths}
source("00-paths.R")
```

Next, we define a function to fetch and parse JSON data from the Firebrowse web site.
```{r fetch}
# get the list of cohorts
library(rjson)
fetch <- function(myurl) {
  stuff <- paste(readLines(myurl), collapse='')
  if (stuff=="") return(NULL)
  struct <- fromJSON(stuff)
  length(struct)
  names(struct)
  gunk <- struct[[1]]
  NR <- length(gunk)
  NC <- length(gunk[[1]])
  daft <- matrix("", nrow=NR, ncol=NC)
  for (i in 1:NR) {
    for (j in 1:NC) {
      x <- gunk[[i]][[j]]
      daft[i,j] <- ifelse(is.null(x), NA, x)
    }
  }
  colnames(daft) <- names(gunk[[1]])
  daft <- as.data.frame(daft)
  daft
}
```

Now we can actually go get the data.
```{r}
myurl <- "http://firebrowse.org/api/v1/Metadata/Cohorts?format=json"
cohorts <- fetch(myurl)
rm(myurl)
cohorts
```
Notice that we get both the short acronym as well as the longer description of
each cancer type.

# Color Schemes
Next, we define a reusable palette of colors to distinguish cancer types in plots.
```{r colors, fig.cap = "Reusable cohort color palette.", fig.width=9}
library(Polychrome)
p36 <- palette36.colors(36)
pal <- createPalette(38, p36[3:35])
dontUse <- c("FPPP", "COADREAD", "GBMLGG", "KIPAN", "STES")
cohortColors <- rep(NA, nrow(cohorts))
cohortColors[!(cohorts$cohort %in% dontUse)] <- pal[1:33]
cohortColors[(cohorts$cohort %in% dontUse)] <- pal[34:38]
names(cohortColors) <- as.character(cohorts$cohort)
swatch(cohortColors)
```

# Case and File Counts
The API at the GDC also allows us to get a list of the numbver of cases and files that
we expect using different technologies. We need t make separate calls for each cohort.
The following function implements the API call.

```{r gdcMeta}
library(rjson)
gdcMeta <- function(COHORT) {
  url <- paste("https://api.gdc.cancer.gov/projects/TCGA-",
               COHORT,
               "?expand=summary,summary.experimental_strategies",
             sep ='')
  stuff <- paste(readLines(url, warn = FALSE), collapse = '')
  thing <- fromJSON(stuff)
  ss <- thing$data$summary
  tot <- data.frame(case_count = ss$case_count,
                    file_count = ss$file_count,
                    experimental_strategy = "All")
  es <- ss$experimental_strategies
  nms <- names(es[[1]])
  extract <- lapply(nms, function(N) { unlist(lapply(es, function(X) { X[N] })) })
  daft <- as.data.frame(extract)
  colnames(daft) <- nms
  daft <- rbind(tot, daft)
  daft$Cohort <- COHORT
  daft
}
```
Now we can fetch all the metadata.
```{r tcgameta}
tumorlist <- as.character(cohorts$cohort)
tumorlist <- tumorlist[!(tumorlist %in% dontUse)]
for (tumor in tumorlist) {
  temp <- gdcMeta(tumor)
  if (exists("tcgaMeta")) {
    tcgaMeta <- rbind(tcgaMeta, temp)
  } else {
    tcgaMeta <- temp
  }
}
tcgaMeta$Cohort <- factor(tcgaMeta$Cohort)
summary(tcgaMeta)
```
Here is an overview of the total number of cases and files.
```{r over}
tcgaMeta[tcgaMeta$experimental_strategy == "All", ]
```

```{r idLists}
ids <- function(COHORT) {
  url <- paste("https://api.gdc.cancer.gov/cases?fields=submitter_id&filters=",
               "%7B%22op%22:%22=%22,%22content%22:%7B%22field%22:%22",
               "project.project_id%22,%22value%22:%5B%22TCGA-",
               COHORT,
               "%22%5D%7D%7D&size=1200",
               sep="")
  stuff <- paste(readLines(url, warn=FALSE), collapse="")
  whatever <- fromJSON(stuff)
  d <- whatever$data$hits
  unlist(lapply(d, function(x) {x$submitter_id}))
}

for (tumor in tumorlist) {
  tempID <- ids(tumor)
  daft <- data.frame(ID = tempID, cancerType = tumor)
  if (exists("idLists")) {
    idLists <- rbind(idLists, daft)
  } else {
    idLists <- daft
  }
}
```

# Finishing Up
Finally, we save the results for reuse in later analyses.
```{r save}
save(cohorts, cohortColors, tcgaMeta, idLists, 
     file = file.path(tcgapaths$clean, "cohorts.Rda"))
```

# Appendix
It's always a good idea to finish up by reporting which versions of the
software tools were used for this analysis:
```{r si}
sessionInfo()
```

