---
title: "TCGA RNA-Seq Data"
author: "Kevin R. Coombes"
date: "8 August 2018"
output: 
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Background
The goal of this report is to collect the normalized RNA-Seq data produced by
The Cancer Genome Atlas (TCGA). Among the possible data sources, we prefer to use
FireBrowse, since it has already collected the data by cancer type.

## Potential data sources

+ [Old TCGA Data Portal](https://tcga-data.nci.nih.gov/docs/publications/tcga/)
+ [cBioPortal](http://www.cbioportal.org/)
+ [FireBrowse](http://firebrowse.org/) -- this one is best for expression data!
+ [Genomic Data Commons](https://gdc-portal.nci.nih.gov/)

# Fetching the Data
We will need an R package to convert JSON-formatted files into R objects.
```{r json}
library(rjson)
```
We also define the paths used to store the data.
```{r paths}
source("00-paths.R")
tcgapaths
```

## URLs for normalized RNA-Seq data, by cancer type
We are going to get the data from the FireBrowse web site. We used their online
"Web API" page to figure out the search URL that will tell us what kinds of RNA
expression data exists for a given cancer type. In this section, we will look at
this list for the "LUSC" cohort.
```{r firebrowse.url}
f <- file.path(tcgapaths$scratch, "firebrowse.Rda")
if (file.exists(f)) {
  load(f)
} else {
  fb <- "http://firebrowse.org/api/v1/Archives/StandardData?format=json&date=2016_01_28"
  misc <- "data_type=mRNASeq&level=3%2C4&page=1&page_size=250&sort_by=cohort"
  url <- paste(fb, misc, "cohort=LUSC", sep="&") # modify here for other cancers
  stuff <- paste(readLines(url), collapse="")
  obj <- fromJSON(stuff)
  obj <- obj[[1]] # this is magic. Not clear why there is an extra top-level one-item list
  rm(fb, misc, url, stuff)
  save(obj, file=f)
}
rm(f)
```

Now we can peek at the first item in the list returned by searching FireBrowse.
```{r obj}
class(obj)
length(obj)
names(obj[[1]])
```
A key distinction between the various kinds of things available is the "protocol".
You can learn more [about the RNA-Seq pipeline here](https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/Expression_mRNA_Pipeline/).
Note that even though I found this page at least once before, it still took google
and almost ten minutes to find it again. Sigh.
```{r protocol}
prot <- unlist(lapply(obj, function(x) x$protocol))
prot
```
We want the "normalized genes" data. So, we pick out the URL for the
correct protocol.
```{r dataurl}
w <- which(prot == "RSEM_genes_normalized") # avoid hard-coding numbers
urlset <- obj[[w]]$urls # RSEM genes normalized
rm(w)
dataurl <- urlset[[1]]
dataurl # note that it is tarred and gzipped
```

Now, most of this URL is the same for every cancer type; the only thing that changes
(in two places) is the "cohort" that describes what kind of cancer we are studying. So,
we are going to create a function that will construct the correct Firebroswse URL
from the cohort name. Specifically, this function creates the URL for the RNA-Seq data, 
in the form of RSEM gene expression counts normalized to the third quartile.
```{r firebrowse}
fb <- function(COHORT) {
  paste("http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/",
    COHORT,
    "/20160128/gdac.broadinstitute.org_",
    COHORT,
    ".Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__",
    "RSEM_genes_normalized__data.Level_3.2016012800.0.0.tar.gz",
    sep="")
}
```
Here is an example:
```{r eg}
x <- fb("COADREAD")
x
nchar(x)
rm(x)
```

## How to fetch the RNA-Seq data
Since the data are in a "_tarred_" and "_gzipped_" file, we have to

1. Download it (using <tt>download.file</tt>).
2. Gunzip it, and untar it (all part of the same <tt>untar</tt> command in R).
3. Delete the <.tar.gz</tt> file (using <tt>unlink</tt>).
4. Read it in two steps, since it includes two header lines, where the
second one is worthless. (See the two <tt>read.table</tt> commands.)
5. Copy the column headers from the first read to the second.
6. Clean up after ourselves (using <t>rm</tt>).
7. Save the results to avoid having to do this again.

All of these steps are performed in the following function:
```{r fetch}
# downlad the file from FireBrowse and store locally
fetch <- function(COHORT) {
  dataurl <- fb(COHORT)
  fname <- basename(dataurl)
  download.file(dataurl, fname)
  cat("uncompressing\n", file=stderr())
  untar(fname, compressed="gzip") # gunzip and untar
  unlink(fname) # remove the tar.gz version
  localdir <- sub(".tar.gz", "", fname)
  file.rename(localdir, COHORT)
  cat("Creating a binary R object\n", file=stderr())
  temp <- file.path(COHORT, dir(COHORT, pattern="RSEM"))
  datasrc <- file.path(COHORT, paste(COHORT, "txt", sep="."))
  file.rename(temp, datasrc)
  datasrc
}

# read the local file
process <- function(LOCALSRC) {
  dumbheads <- read.table(LOCALSRC, sep="\t", header=TRUE, nrows=1, row.names=1)
  ncounts <- read.table(LOCALSRC, sep="\t", header=FALSE, skip=2, row.names=1)
  colnames(ncounts) <- colnames(dumbheads)
  ncounts <- as.matrix(ncounts)
  ncounts
}

# do both
fetchAndProcess <- function(COHORT) {
  localsrc <- fetch(COHORT)
  process(localsrc)
}
```

**WARNING:** _This code will only work in a directory with a **short** name
immediately off the root directory (C:\\) on a Windows machine.  This happens
because, by default, a complete Windows path name cannot exceed 260 characters._

Here is an example of how to use these functions.
```{r example, fig.cap="Histogram of log-transformed data for ACC."}
home <- getwd()
short <- file.path(tcgapaths$raw, "mRNA")
if (!file.exists(short)) dir.create(short)
setwd(short)
f <- "ACC/ACC.txt"
if (!file.exists(f)) {
  ncounts  <- fetchAndProcess("ACC")
} else {
  ncounts <- process(f)
}
setwd(home)
lcounts <- log2(10+ncounts)
hist(lcounts, breaks=123)
```

## Getting all the RNA-Seq data
We already have the names of all the TCGA cohorts.
```{r cohorts}
load(file.path(tcgapaths$clean, "cohorts.Rda"))
cohorts[9:11,]
```
Note that one of these entries does not represent a known type of cancer; instead, it is part
of a pilot study to process samples that were collected and stored as formalin-fixed
paraffin-embedded (FFPE) tissue blocks, We don't want that data.

```{r slurp}
cleanDir <- file.path(tcgapaths$clean, "mRNASeq")
if (!file.exists(cleanDir)) dir.create(cleanDir)
setwd(short)
for (i in 1:nrow(cohorts)) {
  cohort <- as.character(cohorts$cohort)[i]
  if (cohort == "FPPP") { # skip FFPE data
    next
  }
  rawName <- file.path(cohort, paste(cohort, "txt", sep="."))
  cleanName <- file.path(cleanDir, paste(cohort, "Rda", sep="."))
  if (!file.exists(cleanName)) {
    if (!file.exists(rawName)) {
      ncounts <- fetchAndProcess(cohort)
    } else {
      ncounts <- process(rawName)
    }
    lcounts <- log2(10 + ncounts)
    save(ncounts, lcounts, file = cleanName)
  }
}
setwd(home)
```

# Combined data
Here we load the full RNA-Sequencing data set for all 33 TCGA cancer types.
```{r mega}
tumorlist <- as.character(cohorts$cohort)
dontUse <- c("FPPP", "COADREAD", "GBMLGG", "KIPAN", "STES")
tumorlist <- tumorlist[! (tumorlist %in% dontUse)]
f <- file.path(tcgapaths$clean, "mRNAdata.Rda")
if (!file.exists(f)) {
  for (k in 1:length(tumorlist)) {
    tumor <- tumorlist[k]
    cat("working on", tumor, "\n")
    ## load rna-seq data
    load(file.path(tcgapaths$clean, "mRNASeq", paste(tumor, "Rda", sep=".")))
    ## remove unnamed features
    qs <- grep("\\?", rownames(lcounts))
    lcounts <- lcounts[-qs,]
    rm(qs)
    ## Get the preferred genenames by splitting on vertical bar (|) ...
    genename <- unlist(lapply(rownames(lcounts), function(x)  {
      strsplit(x,'[|][[:space:]]*')[[1]][1] 
    }))
    ## remove duplicates names
    dups <- duplicated(genename)
    if (any(dups)) {
      lcounts <- lcounts[!dups,]
      genename <- genename[!dups]
    }
    ## ... and use the names
    rownames(lcounts) <- genename
    lcounts <- lcounts[sort(genename),]
    ct <- rep(tumor, ncol(lcounts))
    if (exists("mRNAdata")) {
      if( any(rownames(mRNAdata) != rownames(lcounts)) ) {
        stop("Cry wolf!")
      }
      mRNAdata <- cbind(mRNAdata, lcounts)
      cancerType <- c(cancerType, ct)
    } else {
      mRNAdata <- lcounts
      cancerType <- ct
    }
    rm(genename, lcounts, ncounts)
    gc()
  } # end for k in 1:length(tumorlist)
  ## Remove 212 features that *always* yield zero counts.
  nzero <- apply(mRNAdata, 1, function(x) sum(x == log2(10)))
  sum(nzero ==ncol(mRNAdata))
  mRNAdata <- mRNAdata[nzero < ncol(mRNAdata),]
  nzero <- nzero[nzero < ncol(mRNAdata)]
  ## Make cancerType a factor
  cancerType <- factor(cancerType)
  save(mRNAdata, nzero, cancerType, tumorlist, file=f)
} # end else
rm(f)
```


# Appendix
It's always a good idea to finish up by reporting which versions of the
software tools were used for this analysis:
```{r si}
sessionInfo()
```

