---
title: "TCGA miR-Seq Data"
author: "Kevin R. Coombes"
date: "8 August 2018"
output: 
  html_document:
    toc: true
    highlight: kate
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
b, strong {color: red; }
i, em {color: blue; }
.defn {color: purple; }
.para {color: purple;
      font-weight: bold;
}
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```

# Background
The goal of this report is to collect the normalized microRNA-Seq data produced by
The Cancer Genome Atlas (TCGA). Among the possible data sources, we prefer to use
FireBrowse, since it has already collected the data by cancer type.

## Potential data sources

+ [Old TCGA Data Portal](https://tcga-data.nci.nih.gov/docs/publications/tcga/)
+ [cBioPortal](http://www.cbioportal.org/)
+ [FireBrowse](http://firebrowse.org/) -- this one is best for expression data!
+ [Genomic Data Commons](https://gdc-portal.nci.nih.gov/)

# Fetching the Data
We will need an R package to convert JSON-formatted files into R objects.
```{r json}
library(rjson)
```
We also define the paths used to store the data.
```{r paths}
source("00-paths.R")
tcgapaths
```

## URLs for normalized miR-Seq data, by cancer type
We are going to get the data from the FireBrowse web site. We used their online
"Web API" page to figure out the search URL that will tell us what kinds of
expression data exist for a given cancer type. In this section, we will look at
this list for the "LUSC" cohort.
```{r firebrowse.url}
f <- file.path(tcgapaths$scratch, "firebrowse-mir.Rda")
if (file.exists(f)) {
  load(f)
} else {
  fb <- "http://firebrowse.org/api/v1/Archives/StandardData?format=json&date=2016_01_28"
  misc <- "data_type=miRSeq&level=3%2C4&page=1&page_size=250&sort_by=cohort"
  url <- paste(fb, misc, "cohort=LUSC", sep="&") # modify here for other cancers
  stuff <- paste(readLines(url), collapse="")
  obj <- fromJSON(stuff)
  obj <- obj[[1]] # this is magic. Not clear why there is an extra top-level one-item list
  rm(fb, misc, url, stuff)
  save(obj, file=f)
}
rm(f)
```

Now we can peek at the first item in the list returned by searching FireBrowse.
```{r obj}
class(obj)
length(obj)
names(obj[[1]])
obj[[1]]$urls
```
The URL that we want is the first one, since this is not just the MD5 hash nor the
MAGE format. Now, most of this URL is the same for every cancer type; the only thing
that changes (in two places) is the "cohort" that describes what kind of cancer we
are studying. So, we are going to create a function that will construct the correct
Firebrowse URL from the cohort name.

```{r spliturl}
oneurl <- obj[[1]]$urls[1]
parts <- strsplit(oneurl, "LUSC")[[1]]
```


```{r firebrowse}
fb <- function(COHORT) {
  paste(parts[1], COHORT,
    parts[2], COHORT,
    parts[3], sep="")
}
```
Here is an example:
```{r eg}
fb("BRCA")
```

## How to fetch the miR-Seq data
Since the data are in a "_tarred_" and "_gzipped_" file, we have to

1. Download it (using <tt>download.file</tt>).
2. Gunzip it, and untar it (all part of the same <tt>untar</tt> command in R).
3. Delete the <.tar.gz</tt> file (using <tt>unlink</tt>).
4. Read it into R.
5. Extract the numeric values.
6. Save the results to avoid having to do this again.

All of these steps are performed in the following functions:
```{r fetch}
# downlad the file from FireBrowse and store locally
fetch <- function(COHORT) {
  dataurl <- fb(COHORT)
  fname <- basename(dataurl)
  val <- try( download.file(dataurl, fname) )
  if(inherits(val, "try-error")) {
    return(NULL)
  }
  cat("uncompressing\n", file=stderr())
  untar(fname, compressed="gzip") # gunzip and untar
  unlink(fname) # remove the tar.gz version
  localdir <- sub(".tar.gz", "", fname)
  file.rename(localdir, COHORT)
  cat("Creating a binary R object\n", file=stderr())
  temp <- file.path(COHORT, dir(COHORT, pattern="RPM.txt"))
  datasrc <- file.path(COHORT, paste(COHORT, "txt", sep="."))
  file.rename(temp, datasrc)
  datasrc
}
# read the local file
process <- function(LOCALSRC) {
  ncounts <- read.table(LOCALSRC, sep="\t", header=TRUE, row.names = 1)
  ncounts <- as.matrix(ncounts)
  ncounts[is.na(ncounts)] <- 0 # really?
  ncounts
}
# do both
fetchAndProcess <- function(COHORT) {
  localsrc <- fetch(COHORT)
  if(is.null(localsrc)) return(NULL)
  process(localsrc)
}
```

**WARNING:** _This code will only work in a directory with a **short** name
immediately off the root directory (C:\\) on a Windows machine.  This happens
because, by default, a complete Windows path name cannot exceed 260 characters._

Here is an example of how to use these functions.
```{r example, fig.cap="Histogram of log-transformed data for ACC."}
home <- getwd()
short <- file.path(tcgapaths$raw, "miRSeq")
if (!file.exists(short)) dir.create(short)
setwd(short)
f <- "ACC/ACC.txt"
if (!file.exists(f)) {
  ncounts  <- fetchAndProcess("ACC")
} else {
  ncounts <- process(f)
}
setwd(home)
lcounts <- log2(1 + ncounts)
hist(lcounts, breaks=123)
```

## Getting all the miR-Seq data
We already have the names of all the TCGA cohorts.
```{r cohorts}
load(file.path(tcgapaths$clean, "cohorts.Rda"))
cohorts[9:11,]
```
Note that one of these does not represent a known type of cancer; instead, it is part
of a pilot study to process samples that were collected and stored as formalin-fixed
paraffin-embedded (FFPE) tissue blocks, We don't want that data.

```{r slurp}
cleanDir <- file.path(tcgapaths$clean, "miRSeq")
if (!file.exists(cleanDir)) dir.create(cleanDir)
setwd(short)
for (i in 1:nrow(cohorts)) {
  cohort <- as.character(cohorts$cohort)[i]
  if (cohort %in% c("FPPP", "LAML")) { # skip FFPE data; there is none for LAML
    next
  }
  rawName <- file.path(cohort, paste(cohort, "txt", sep="."))
  cleanName <- file.path(cleanDir, paste(cohort, "Rda", sep="."))
  if (!file.exists(cleanName)) {
    if (!file.exists(rawName)) {
      ncounts <- fetchAndProcess(cohort)
    } else {
      ncounts <- process(rawName)
    }
    if(!is.null(ncounts)) {
      lcounts <- log2(1 + ncounts)
      save(ncounts, lcounts, file = cleanName)
    }
  }
}
setwd(home)
```

# Combined data
Here we load the full miR-Sequencing data set for all 32(!) TCGA cancer types.
```{r mega}
tumorlist <- as.character(cohorts$cohort)
dontUse <- c("FPPP", "COADREAD", "GBMLGG", "KIPAN", "STES", "LAML") # No LAML miR data
tumorlist <- tumorlist[! (tumorlist %in% dontUse)]
f <- file.path(tcgapaths$clean, "mirData.Rda")
if (!file.exists(f)) {
  for (k in 1:length(tumorlist)) {
    tumor <- tumorlist[k]
    cat("working on", tumor, "\n")
    ## load miR-seq data
    load(file.path(tcgapaths$clean, "miRSeq", paste(tumor, "Rda", sep=".")))
    ## Get the preferred genenames by splitting on vertical bar (|) ...
    genename <- unlist(lapply(rownames(lcounts), function(x)  {
      strsplit(x,'[|][[:space:]]*')[[1]][1] 
    }))
    ## remove duplicates names
    dups <- duplicated(genename)
    if (any(dups)) {
      lcounts <- lcounts[!dups,]
      genename <- genename[!dups]
    }
    ## ... and use the names
    rownames(lcounts) <- genename
    lcounts <- lcounts[sort(genename),]
    ct <- rep(tumor, ncol(lcounts))
    if (exists("miRdata")) {
      miRdata <- cbind(miRdata, lcounts)
      cancerType <- c(cancerType, ct)
    } else {
      miRdata <- lcounts
      cancerType <- ct
    }
    rm(genename, lcounts, ncounts)
    gc()
  } # end for k in 1:length(tumorlist)
  ## Make cancerType a factor
  cancerType <- factor(cancerType)
  ## Remove features that *always* yield zero counts.
  MN <- min(miRdata)
  nzero <- apply(miRdata, 1, function(x) sum(x == MN))
  miRdata <- miRdata[nzero < ncol(miRdata),]
  nzero <- nzero[nzero < ncol(miRdata)]
  save(miRdata, nzero, cancerType, tumorlist, file=f)
} # end else
rm(f)
```

```{r chi, echo=FALSE,eval=FALSE}
chis <- t(apply(miRdata, 1, function(x) {
  chi <- chisq.test(table(x, cancerType))
  c(chi$statistic, chi$p.value)
}))
summary(chis)
```

# Appendix
It's always a good idea to finish up by reporting which versions of the
software tools were used for this analysis:
```{r si}
sessionInfo()
```

